---
title: "Simulating the Central Limit Theorem Using the Exponential Eistribution"
author: "Alexander Sickert"
date: "27 September 2015"
output: html_document
---


   
#Overview: 

This project investigates several features of the Central Limit Theorem using the exponential distribution. The exponential distribution is being applied using lambda = 0.2.

#Simulations: 

##Sample Mean versus Theoretical Mean: 

The mean of exponential distribution is 1/lambda. With a lambda of 0.2 the theoretical mean is 5.

The following code takes 1000 averages of 40 exponentials. The sample growths from 1 to 1000 and for each step the mean is being calculated. 


```{r}
n <- 1000; 
means = c()
meanValueCumulative = 0
for (i in 1 : n){
  meanValue = mean(rexp(40, 0.2))
  meanValueCumulative = meanValueCumulative + meanValue
  meanNow = meanValueCumulative / i 
  means = c(means, meanNow)
} 

plot(1 : n, means, type = "l", lwd = 2, frame = FALSE, ylab = "cumulative means", xlab = "sample size") 

abline(h = 5)
```

It can be observed that the mean is increasingly moving towards the theoretical mean of 5 

##Sample Variance versus Theoretical Variance 

The variance of exponential distribution is 1/lambda^2. With a lambda of 0.2 the theoretical variance is 25.

The following code takes 1000 averages of 40 exponentials. The sample growths from 1 to 1000 and for each step the variance is being calculated. 

```{r}
n <- 1000; 
means = c()
meanValueCumulative = 0
for (i in 1 : n){
  meanValue = var(rexp(40, 0.2))
  meanValueCumulative = meanValueCumulative + meanValue
  meanNow = meanValueCumulative / i 
  means = c(means, meanNow)
} 

plot(1 : n, means, type = "l", lwd = 2, frame = FALSE, ylab = "cumulative variance", xlab = "sample size") 

abline(h = 25)
```

It can be observed that the variance is increasingly moving towards the theoretical variance of 25. 

##Distribution

The following code takes 1000 averages of 40 exponentials. Then a histogram is being created. In addition a normal curve is being displayed in red color.  

```{r}
mns = NULL
for (i in 1 : 1000) mns = c(mns, mean(rexp(40, 0.2)-5))
hist(mns, freq = FALSE, col = "grey")
curve(dnorm, col = 2, add = TRUE)
```

It can be observed tha the shape of the histogram is similar to a normal distribution. This effect becomes more visible if we as an alternative use 1000 exponentials and create a histogram of them. The resutling historgram has exponential shape and not normal shape.  

```{r}
vals = rexp(1000, 0.2)
hist(vals, freq = FALSE, col = "grey")
```
