---
title: "Simulating the Central Limit Theorem Using the Exponential Eistribution"
author: "Alexander Sickert"
date: "23rd September 2015"
output: html_document
---


   
#Overview: 

In a few (2-3) sentences explain what is going to be reported on.

In this project you will investigate the exponential distribution in R and compare it with the Central Limit Theorem.

 The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations. You will investigate the distribution of averages of 40 exponentials. Note that you will need to do a thousand simulations.
 
 Illustrate via simulation and associated explanatory text the properties of the distribution of the mean of 40 exponentials.  You should
1. Show the sample mean and compare it to the theoretical mean of the distribution.
2. Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
3. Show that the distribution is approximately normal.

In point 3, focus on the difference between the distribution of a large collection of random exponentials and the distribution of a large collection of averages of 40 exponentials. 

As a motivating example, compare the distribution of 1000 random uniforms

hist(runif(1000))

and the distribution of 1000 averages of 40 random uniforms
```{r}
mns = NULL
for (i in 1 : 1000) mns = c(mns, mean(runif(40)))
hist(mns)
```

This distribution looks far more Gaussian than the original uniform distribution!


This exercise is asking you to use your knowledge of the theory given in class to relate the two distributions.  
Confused?  Try re-watching video lecture 07 for a starter on how to complete this project.

#Simulations: 

##Sample Mean versus Theoretical Mean: 

The mean of exponential distribution is 1/lambda. With a lambda of 0.2 the theoretical mean is 5.

```{r}
n <- 1000; 
means = c()
meanValueCumulative = 0
for (i in 1 : n){
  meanValue = mean(rexp(40, 0.2))
  meanValueCumulative = meanValueCumulative + meanValue
  meanNow = meanValueCumulative / i 
  means = c(means, meanNow)
} 

plot(1 : n, means, type = "l", lwd = 2, frame = FALSE, ylab = "cumulative means", xlab = "sample size") 

abline(h = 5)
```



##Sample Variance versus Theoretical Variance: 

The variance of exponential distribution is 1/lambda^2. With a lambda of 0.2 the theoretical variance is 25.



```{r}
n <- 1000; 
means = c()
meanValueCumulative = 0
for (i in 1 : n){
  meanValue = var(rexp(40, 0.2))
  meanValueCumulative = meanValueCumulative + meanValue
  meanNow = meanValueCumulative / i 
  means = c(means, meanNow)
} 

plot(1 : n, means, type = "l", lwd = 2, frame = FALSE, ylab = "cumulative means", xlab = "sample size") 

abline(h = 25)
```

deviation is also 1/lambda

##Distribution: Via figures and text, explain how one can tell the distribution is approximately normal. 



```{r}
mns = NULL
for (i in 1 : 1000) mns = c(mns, mean(rexp(40, 0.2)-5))
hist(mns, freq = FALSE, col = "grey")
curve(dnorm, col = 2, add = TRUE)
```



