---
title: "Weather Events and its Damages - Hurricane, Heat and Flood biggest problem"
author: "Alexander Sickert"
date: "21 July 2015"
output: html_document
---

# Synopsis

The following analysis is built on the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database which covers a time from from 1950 to 2011. 
The goal of the analysis is to highlight the events that are most harmful for man and economically most damaging. 

The results are

- Flood and Hurricane/Tornado events are economically most devastating
- Looking into trends both event types are increasing in quantity over time. 
- Tornados and excessive head cause most fatalities

As a result I recommend that government takes more measures to prevent fatalities during heat periods. I also recommend investigating how the damages can be reduced for example by not allowing houses being built in risky areas.




# Data Processing

Loading raw data
```{r, cache=TRUE }
dataRaw = read.csv("repdata_data_StormData.csv.bz2", sep = ",", header = TRUE, as.is = TRUE)
```

###Clean data - quantifies of economic damage

To be able to aggregate data about economic damage the characters need be converted into numbers and multiplied with the decimal value of damage.
```{r, echo=FALSE}

dataProcessing <- transform(dataRaw, PROPDMGEXP = toupper(gsub("\\s", "", PROPDMGEXP) ))
dataProcessing <- transform(dataProcessing, CROPDMGEXP = toupper(gsub("\\s", "", CROPDMGEXP) ))

dataProcessing <- transform(dataProcessing, PROPDMG = ifelse(PROPDMGEXP == "H", PROPDMG * 100, ifelse(PROPDMGEXP == "K", PROPDMG * 1000, ifelse(PROPDMGEXP == "M", PROPDMG * 1000000, ifelse(PROPDMGEXP == "B", PROPDMG * 1000000000, PROPDMG)))))

dataProcessing <- transform(dataProcessing, CROPDMG = ifelse(CROPDMGEXP == "H", CROPDMG * 100, ifelse(CROPDMGEXP == "K", CROPDMG * 1000, ifelse(CROPDMGEXP == "M", CROPDMG * 1000000, ifelse(CROPDMGEXP == "B", CROPDMG * 1000000000, CROPDMG)))))

# calculating total damage = property damange + crop damange
dataProcessing$DAMAGE = dataProcessing$CROPDMG + dataProcessing$PROPDMG
```

Adding an additional column that holds the year of the event. It is extraced from the date of the event. 

```{r, echo=FALSE}

# adding a year for later comparison based on years

dataProcessing$YEAR = as.numeric(format(as.Date(dataProcessing$BGN_DATE, "%m/%d/%Y"), "%Y"))
```




# Analysis

###Aggregating  data of economic damage

The aim is to find out which event causes the biggest damage, how many events occur and what the average damage of an event is. 

```{r}
library(sqldf) 
dataEconomic = sqldf("select EVTYPE, avg(DAMAGE), count(DAMAGE), sum(DAMAGE)  from dataProcessing group by
EVTYPE order by sum(DAMAGE)  desc ") 

# print(dataEconomic)

```

###Extract the top 10 event types

As the list of events is very long we take into consideration only the top ten events. 
```{r}
dataEconomicTop = dataEconomic[1:10,c("EVTYPE", "avg(DAMAGE)", "count(DAMAGE)", "sum(DAMAGE)")]
colnames(dataEconomicTop) <- c("EVTYPE", "AVERAGE", "COUNT", "SUM")
print(dataEconomicTop)
```


###Development of damaging events over time

As a next step we analyze how the biggest event types develop over time to see if there are trends 

```{r}

# FLOOD

dataEconomicByYearFlood = sqldf("select YEAR, count(*) as EVENTS from dataProcessing where trim(EVTYPE) in ( 'FLOOD' ) and DAMAGE > 0  group by YEAR order by YEAR  asc ") 

# print(dataEconomicByYearFlood)

plot(dataEconomicByYearFlood$YEAR, dataEconomicByYearFlood$EVENTS,  type="l",  xlab="Year", ylab="# of Events", ylim=c(0,3000), col="black")

# HURRICANE/TYPHOON', "TORNADO

dataEconomicByYearHurricane = sqldf("select YEAR, count(*) as EVENTS from dataProcessing where trim(EVTYPE) in ( 'HURRICANE/TYPHOON', 'TORNADO', 'HURRICANE' ) and DAMAGE > 0  group by
YEAR order by YEAR  asc ") 

# print(dataEconomicByYearHurricane)

lines(dataEconomicByYearHurricane$YEAR, dataEconomicByYearHurricane$EVENTS, type="l", pch=22, lty=1, col="red")


# STORM SURGE

dataEconomicByYearSurge = sqldf("select YEAR, count(*) as EVENTS   from dataProcessing where trim(EVTYPE) in ( 'STORM SURGE' ) and DAMAGE > 0  group by
YEAR order by YEAR  asc ")

# print(dataEconomicByYearSurge)

lines(dataEconomicByYearSurge$YEAR, dataEconomicByYearSurge$EVENTS, type="l", pch=22, lty=1, col="green")

# HAIL

dataEconomicByYearHail = sqldf("select YEAR, count(*) as EVENTS   from dataProcessing where trim(EVTYPE) in ( 'HAIL' ) and DAMAGE > 0  group by
YEAR order by YEAR  asc ") 

# print(dataEconomicByYearHail)

lines(dataEconomicByYearHail$YEAR, dataEconomicByYearHail$EVENTS, type="l", pch=22, lty=1, col="blue")

legend(1995,2900, c('Flood','Hurricane','Surge','Hail'),  lty=c(1,1,1,1), col=c('black','red' ,'green','blue') ) 

title("# of Economic Damage Events Over Time")


```

##aggregate data of fatalities

The aim is to find out which event causes the biggest number of fatalities, how many events occur and what the average number of fatalities of an event is. 

```{r}
dataFatalities = sqldf("select EVTYPE, avg(FATALITIES), count(FATALITIES), sum(FATALITIES)  from dataProcessing group by
EVTYPE order by sum(FATALITIES)  desc ") 

# print(dataFatalities)

```

###Extract top 10

As there are many different events the analysis focuses on the ten event types with biggest total number of fatalities. 

```{r}
dataFatalitiesTop = dataFatalities[1:10,c("EVTYPE", "avg(FATALITIES)", "count(FATALITIES)", "sum(FATALITIES)")]
colnames(dataFatalitiesTop) <- c("Event Type", "Avg. Fatalities per Event", "# Events", "SUM of Fatalities")
print(dataFatalitiesTop)

```

###Display development of events over time

To see if there are trends over time the event types that cause most fatalities are displayed over time. Several events that fall under the term "wind" can be grouped together. 

```{r}
# 'TORNADO' , 'HIGH WIND', 'TSTM WIND'

dataFatalitiesByYearWind = sqldf("select YEAR, count(*) as EVENTS from dataProcessing where trim(EVTYPE) in ('TORNADO' , 'HIGH WIND', 'TSTM WIND' ) and FATALITIES > 0  group by YEAR order by YEAR  asc ") 

print(dataFatalitiesByYearWind)

plot(dataFatalitiesByYearWind$YEAR, dataFatalitiesByYearWind$EVENTS,  type="l",  xlab="Year", ylab="# of Events", ylim=c(0,100), col="black")

# 'EXCESSIVE HEAT', 'HEAT'

dataFatalitiesByYearHeat = sqldf("select YEAR, count(*) as EVENTS from dataProcessing where trim(EVTYPE) in ( 'EXCESSIVE HEAT', 'HEAT' ) and FATALITIES > 0  group by YEAR order by YEAR  asc ") 

print(dataFatalitiesByYearHeat)

lines(dataFatalitiesByYearHeat$YEAR, dataFatalitiesByYearHeat$EVENTS, type="l", pch=22, lty=1, col="red")

# make lines and legend


legend(1950,95, c('Wind','Heat'),  lty=c(1,1,1), col=c('black','red') ) 

title("# of Fatalities of Heat and Wind Events Over Time")


```


###Display Flood events


```{r}

# FLASH FLOOD  FLOOD 

dataFatalitiesByYearFlood = sqldf("select YEAR, count(*) as EVENTS from dataProcessing where trim(EVTYPE) in ( 'FLASH FLOOD',  'FLOOD'  ) and FATALITIES > 0  group by YEAR order by YEAR  asc ") 

print(dataFatalitiesByYearFlood)

plot(dataFatalitiesByYearFlood$YEAR, dataFatalitiesByYearFlood$EVENTS,  type="l",  xlab="Year", ylab="# of Events", ylim=c(0,250), col="black")

title("# of Flood Events Over Time")

```

# Results 

### Restults on Economic Damage

- Flood and Wind related events are most damaging. 
- Damages from surge sopped after 2005. This might be a data collection error. 
- Damages from wind related events and flood are getting more since 2005


### Restults on Fatalities

- Tornado, Head and Flood events cause the most fatalities
- Heat events are shown in the database only since 1994. This can distort the analaysis. 
- Flood events show an increasing trend of occurence. 

