---
title: "Tooth Growth vs. Supplements VC and OJ"
author: "Alexander Sickert"
date: "26 September 2015"
output: html_document
---

#Load Data

First the data is being loaded. Then, using the str() function a first high level descriptionon of the data is being obtained. 

```{r}
data(ToothGrowth)
str(ToothGrowth)
```

We learn that the data has 3 variables. It could be that there are N.A. values and we test if this is the case.  

```{r}
sum(is.na(ToothGrowth$len))
sum(is.na(ToothGrowth$dose))
sum(is.na(ToothGrowth$supp))
```

As a result we can say that the dataset
- has 3 variables
- 6 datasets
- the "supp" valirable has two levels
- the "dose" and "len" variables hare numeric
- there are no N.A. values

# Summary of Data

From the previous section we learned that the dataset contains dat for two different supplements. To understand the data better the two supplements are handled separately. 

Data for supplement "VC"

To create a boxplot diagram the vairable "dosis" is being converted into a factor/level. 

```{r}
VC = subset(ToothGrowth, supp == "VC" , c(len, dose))
VC = as.data.frame(VC)
VC$dose <- factor(VC$dose)
boxplot(len ~ dose, data = VC, main="Dosis of VC vs. Length", xlab="Dosis VC", ylab="Length", color="red")
```



Data for supplement "OJ"

To create a boxplot diagram the vairable "dosis" is being converted into a factor/level. 

```{r}
OJ = subset(ToothGrowth, supp == "OJ" , c(len, dose))
OJ = as.data.frame(OJ)
OJ$dose <- factor(OJ$dose)
boxplot(len ~ dose, data = OJ, main="Dosis of OJ vs. Length", xlab="Dosis OJ", ylab="Length", color="red")
```

Comparing the two graphs OJ and VC indicates that observations in OJ are more spread out for a lower dosis. Bit graphs tell that the mean value of tooth growth increases with an increase of the dosis of the supplement. 


#Tooth Growth by Supplement and Dose

Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)

In order to understand if an increase of dosis has an impact on tooth growth the t-test is applied. As the t-test can only compare two sets of observations the dosis 0.5 and 2 was chosen. 

The Null-hypothesis for this test is: There is no difference in tooth growth between the tow sets of observations. 

T-test for supplement OJ
```{r}
OJt = subset(OJ, dose == c(0.5,2) , c(len, dose))
t.test(len ~ dose, paired = FALSE, val.equal=TRUE, data = OJt)
```

T-test for supplement VC
```{r}
VCt = subset(VC, dose == c(0.5,2) , c(len, dose))
t.test(len ~ dose, paired = FALSE, val.equal=TRUE, data = VCt)
```

Another approach is to understand if the supplements have the same effect or if one of the supplements is more effective than the other. 

```{r}
VCOJt <- subset(ToothGrowth,select=c(len, supp)) 
t.test(len ~ supp, paired = FALSE, val.equal=TRUE, data = VCOJt)
```


#Conclusions and Assumptions

##Conclusion

The t-test for supplement VC confirms what can be observed but the boxplot diagram: The null hypothesis must be rejhected due to the very low p-value in favour of the alternative hapothesis that there is an increase tooth growth depending on the dosis. 

The same conclusion can be made for the OJ supplement. 

T-test for comparison of VC vs. OJ: The null hypothesis is that there is no difference between the impact of the two supplements. The high p-value indicates that null hyposesis cannot be rejected and there is no significant difference between the impact of VC and OJ on tooth growth. 

##Assumptions

Welch's t-test assumes that the observations are normally distributed. The analysis did not verify this. 

Another assumption is that there are no other factors than the supplements which impacted the tooth growth. Correlation does not imply causation. 



