---
title: "Regression Models - Course Project MTCARS"
author: "Alexander Sickert"
date: "22 November 2015"
output: 
  html_document: 
    fig_height: 4
    fig_width: 6
---


## Executive Summary

The aim of this analysis is to explore the relationship between a set of variables and miles per gallon (MPG) (outcome). Two central questions should be answered:

a) Is an automatic or manual transmission better for MPG?
b) Quantify the MPG difference between automatic and manual transmissions

The dataset used for the analysis was extracted from the 1974 Motor Trend US magazine.

The result of the analysis:

a) On average a manual transmission is better for MPG
b) If we ignore all other parameters of the data, manual transmission has ***7.3*** higher MPG wiht a 95% confidence interval of ***3.2*** to ***11.3***. However, if we use a model that includes 3 predictors, then the automatic/manual predictor accounts for ***2.94*** increase in MPG when moving from automatic to manual with a 95% confidence interval of ***0.05 - 5.83*** mpg.

## Exploratory Analysis

First we have a high level look into the dataset. See  ***Appendix: First Look. *** 

The datasets consists of 11 variables and does  not have incomplete records. 

A simple bos plot provides a first look at the difference in MPG depending on Automatic and Manual. It indicates a difference in MPG. See  ***Appendix: Boxplot ***

Based on this finding we can quantify the differences using a t-test. See  ***Appendix: T-Test.  ***

**We learn that on average an automatic transmission consumes more fuel than a manual transmission (17.1 vs. 24.4 mpg). The 95% confidence interval of the difference is 3.2 - 11.3.  **

It is desirable that the outcome (MPG) is normal distrubuted and hence we check it with a histogram, see Appendix:Histogram. According to the histogram the data is not exactly normal distributed but we assume it is sufficiently similar to a normal distribution. 

## Model Selection

From the exploratory analysis we could conclude a simple linear regression using Automatic/Manual as the predictor could be sufficient. ***See Appendix: Simple Model***

What we learn is this: As the P-value is low, we can reject the null hypothesis that there is no difference between Manual and Automatic. However, the parameter explains only  ***34% *** of the variance that we can observe in mpg. Therefore we want to find a better model. 

We could do a manual proeces and start with a model that includes all varables and then and step by step eliminate the variables that do not have significant impact on the outcome. A starting point for such a process is the list of coefficients, See  ***Appendix:Coefficients ***

For example the parameters cyl and carb have low impact on the outcome and a high p-value and we could construct a new model. However, there exists in R a function that does all that automatically. See  ***Appendix: Stepwise *** 

This model contains only three predictors: wt, qsec and am. The model is able to explain  ***83% *** of the variance in mpg. We therefore consider this model a good candidate for further diagnostics. 


## Further Exploratory Analysis and Diagnostics 

To see if our model appears reasonable we plot all 3 parameters  against each other. See  ***Appendix: Pairs ***


If the model is suitable, then we should not find particular  patterns when we carry out a residual plot. See  ***Appendix: Residual Plot ***

We cannot observe a pattern that would indicate that the model is not suitable. 

Finally we want to quantify the findings. We learn from the summary that the thee is an estimated increase in mpg of **2.94** when moving from automatic to manual. See  ***Appendix: Summary & Coefficients ***

The **95%** confidence interval for this finding is **0.05 - 5.83**. 

## Appendix

### First look

```{r}
data("mtcars")
str(mtcars)
```

### Boxplot

```{r}
mtcarsMod = mtcars
mtcarsMod$am <- as.factor(mtcarsMod$am)
levels(mtcarsMod$am) <- c("Automatic","Manual") 
boxplot(mpg ~ am, data = mtcarsMod, ylab = "MPG", main = "Miles per Gallon vs. Transmission Type")
```

### T-Test
```{r}
mtcarsAutomatic <- mtcarsMod[mtcarsMod$am == "Automatic", ] 
mtcarsManual = mtcarsMod[mtcarsMod$am == "Manual", ] 
t.test(mtcarsAutomatic$mpg, mtcarsManual$mpg)
average = 24.4 - 17.1
average
```

### Histogram

```{r}
hist(mtcars$mpg, breaks = 10, xlab = "MPG", main = "Histogram MPG")
```

### Simple Model 


```{r}
fit = lm(mpg ~ am, data = mtcarsMod)
summary(fit)
```

### Coefficients

```{r}
summary(lm(mpg ~ ., data = mtcars))$coefficients
```

### Stepwise

```{r}
stepwise = step(lm(data = mtcarsMod, mpg ~ .), trace = 0 )
summary(stepwise)
```

### Pairs

```{r}
mtcarsMod = mtcars
mtcarsMod$am <- as.factor(mtcarsMod$am)
levels(mtcarsMod$am) <- c("Automatic","Manual") 
pairs(mtcarsMod[, c("mpg", "wt", "qsec", "am")], panel = panel.smooth, main = "Relationship of all 4 paramters")
```

### Residual Plot

```{r}
optiModel = lm(mpg ~ am + wt + qsec, data = mtcarsMod)
par(mfrow=c(2,2))
plot(optiModel)
```

### Summary & Coefficients

```{r}
summary(optiModel)
```

```{r}
coef = summary(optiModel)$coefficients
estimate = coef["amManual", 1] 
interval = estimate + c(-1, 1) * qt(0.975, df = optiModel$df) * coef["amManual", 2]
interval
```

